---
title: "Using single-cell chromatin accessibility sequencing to characterize CD4+
  T cells from murine tissues - downstream analysis\n"
author: |
  Kathrin Luise Braband, Annekathrin Silvia Nedwed, Sara Salome Helbich, Malte Simon, Niklas Beumer, Benedikt Brors, Federico Marini and Michael Delacher
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    number_sections: yes
  html_document:
    toc: yes
    toc_float: yes
    code_folding: show
    number_sections: yes
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(ArchR)
library(tidyverse)
library(dplyr)
library(pheatmap)
library(cowplot)
library(SingleR)
library(celldex)
library(scRNAseq)
library(biomaRt)
library(scuttle)
library(BSgenome.Mmusculus.UCSC.mm10)
library(knitr)
```

```{r setInitialParameters}
# Set seeds and initial parameters for ArchR
RNGkind("L'Ecuyer-CMRG")
set.seed(1)

# Number of cores to use (adapt to your machine, but has to be > 1 otherwise 
# plotTSSenrichment does not work properly)
addArchRThreads(threads = 8)

# Reference genome (adapt to your data, use the same reference that was used for 
# the alignment)
addArchRGenome("mm10")
```

```{r load ArchRProject proj, warning=FALSE, message=FALSE}
# Load ArchRProject
proj = loadArchRProject(
  path = "ArchRProject_final", 
  force = TRUE, 
  showLogo = FALSE
  )
```

# Final filtered dataset
## Visualisation in UMAP embedding
```{r final Dataset - visualization in UMAP embedding, message=FALSE, warning=FALSE}
# Color by clusters
p_clusters = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "Clusters", 
  embedding = "UMAP", 
  size = 0.5
  )

# Color by samples
p_samples = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "Sample", 
  embedding = "UMAP", 
  size = 0.5
  )

plot_grid(p_clusters, p_samples, align = "h", ncol = 2)

# Color by doublet enrichment score
p_doubscore = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "DoubletScore", 
  embedding = "UMAP", 
  plotAs = "points", 
  size = 0.5
)

p_doubenr=plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "DoubletEnrichment", 
  embedding = "UMAP", 
  plotAs = "points", 
  size = 0.5
)

# Color by number of fragments per cell
p_nfrags = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "nFrags", 
  embedding = "UMAP", 
  plotAs = "points", 
  size = 0.5
)

# Plot
plot_grid(p_doubscore, p_doubenr, p_nfrags, align = "h", ncol = 3, scale = 1.1)
```

```{r save UMAPs, warning=FALSE, message=FALSE, echo=FALSE}
# Save as PDF
plotPDF(p_clusters, p_samples, p_doubscore, p_doubenr, p_nfrags, 
  name = "UMAP_clusters_samples_doubscore_nfrags.pdf", 
  ArchRProj = proj, 
  addDOC = FALSE, 
  width = 5, 
  height = 5
)
```

# Cell type annotation
## Manual cluster-based annotation using prior-knowledge marker genes
```{r magic, warning=FALSE, message=FALSE}
# Overlay gene scores on UMAP embedding, use MAGIC smoothing
# Get marker features
markersGS = getMarkerFeatures(
  ArchRProj = proj, 
  useMatrix = "GeneScoreMatrix", 
  groupBy = "Clusters",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon"
)

# Define which genes to plot
markerGenes = c("Cd4", "Il2ra", "Foxp3", "Batf", "Ccr8", "Ikzf2", "Rorc", "Tbx21", 
  "Ifng", "Gata3", "Il2", "Sell")

# Add impute weights
proj = addImputeWeights(proj)

# Plot
magic_genes = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "GeneScoreMatrix", 
  name = markerGenes, 
  embedding = "UMAP",
  plotAs = "points",
  imputeWeights = getImputeWeights(proj),
  size = 0.5
)

# Plot all genes in cowplot
magic_genes_cow = lapply(magic_genes, function(x){
  x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
      axis.text.x=element_blank(), 
      axis.ticks.x=element_blank(), 
      axis.text.y=element_blank(), 
      axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 4), magic_genes_cow))

# Save as PDF
plotPDF(magic_genes_cow,  
  name = "UMAP_genescores.pdf", 
  ArchRProj = proj, 
  addDOC = FALSE, 
  width = 5, 
  height = 5
)
```

```{r markers as violin plots, message=FALSE, warning=FALSE}
p_Cd4 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Cd4",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Cd4

p_Il2ra = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Il2ra",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Il2ra

p_Foxp3 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Foxp3",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Foxp3

p_Batf = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Batf",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Batf

p_Ccr8 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Ccr8",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Ccr8

p_Ikzf2 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Ikzf2",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Ikzf2

p_Rorc = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Rorc",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Rorc

p_Tbx21 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Tbx21",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Tbx21

p_Ifng = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Ifng",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Ifng

p_Gata3 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Gata3",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Gata3

p_Klrg1 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Klrg1",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Klrg1

p_Il2 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Il2",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Il2
```

```{r final project: save violin plots, warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE}
# Save as PDF
plotPDF(p_Cd4, p_Il2ra, p_Foxp3, p_Batf, p_Ccr8, p_Ikzf2, p_Rorc, p_Tbx21, p_Ifng, p_Gata3, p_Klrg1, p_Il2, 
  name = "UMAP_final_violin.pdf", 
  ArchRProj = proj, 
  addDOC = FALSE, 
  width = 5, 
  height = 5
)
```

```{r magic Tregs, message=FALSE, warning=FALSE}
# Look at tissue Tregs and their precursor stages in the spleen
markerGenes_treg = c("Foxp3", "Batf", "Nfil3", "Klrg1", "Gata3", "Il1rl1", "Id3", 
  "Id2", "Ikzf2")

# Plot
magic_genes_treg = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "GeneScoreMatrix", 
  name = markerGenes_treg, 
  embedding = "UMAP",
  plotAs = "points", 
  imputeWeights = getImputeWeights(proj),
  size = 0.5
)

# Plot all genes in cowplot
magic_genes_cow_treg = lapply(magic_genes_treg, function(x){
  x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
      axis.text.x=element_blank(), 
      axis.ticks.x=element_blank(), 
      axis.text.y=element_blank(), 
      axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3), magic_genes_cow_treg))
```

```{r save magic UMAPs Treg, warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE}
# Save
plotPDF(plotList = magic_genes_treg, 
        name = "Plot-UMAP_MAGIC-Treg-Genes_proj.pdf", 
        ArchRProj = proj, 
        addDOC = FALSE, width = 5, height = 5)
```

```{r treg markers as violin plots, message=FALSE, warning=FALSE}
# To further characterize Treg cells in our dataset, we look at various Treg
# markers
p_Foxp3 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Foxp3",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Foxp3

p_Batf = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Batf",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Batf

p_Nfil3 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Nfil3",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Nfil3

p_Klrg1 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Klrg1",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Klrg1

p_Gata3 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Gata3",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Gata3

p_Il1rl1 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Il1rl1",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Il1rl1

p_Id3 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Id3",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Id3

p_Id2 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Id2",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Id2

p_Ikzf2 = plotGroups(
  ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "GeneScoreMatrix", 
  name = "Ikzf2",
  plotAs = "violin",
  alpha = 0.4,
  addBoxPlot = TRUE
)
p_Ikzf2
```

```{r final project treg: save violin plots, warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE}
# Save as PDF
plotPDF(p_Foxp3, p_Batf, p_Nfil3, p_Klrg1, p_Gata3, p_Il1rl1, p_Id3, p_Id2, p_Ikzf2, 
  name = "UMAP_final_violin_treg_genes.pdf", 
  ArchRProj = proj, 
  addDOC = FALSE, 
  width = 5, 
  height = 5
)
```

```{r define clusters, message=FALSE, warning=FALSE}
# Define clusters for comparisons/ project subsetting (to be performed later)
tisTreg_cluster = c("C8", "C12", "C14", "C15") #C8 spleen, C15 fat, C14 skin, C12 colon
tTreg_cluster = c("C10", "C11") #C11 early precursor, C10 late precursor  
pTreg_cluster = "C13"

# Also scATAC_8 is from spleen, however this sample includes only CD25+ cells 
# whereas all other samples are CD4+
spleen = "scATAC_1" 
skin = "scATAC_9"
colon = "scATAC_4"
fat = "scATAC_5"
tissue = c(skin, colon, fat)
```

```{r rename clusters accordingly, message=FALSE, warning=FALSE, eval=FALSE}
# Add cluster names to AchRProject
clusters = as.data.frame(proj@cellColData$Clusters)
colnames(clusters) = "V1"

clusters = clusters %>% mutate(across("V1", str_replace, "^C8$", "C8 tisTreg spleen")) 
clusters = clusters %>% mutate(across("V1", str_replace, "^C15$", "C15 tisTreg fat"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C14$", "C14 tisTreg skin"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C12$", "C12 tisTreg colon"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C13$", "C13 pTreg colon")) 
clusters = clusters %>% mutate(across("V1", str_replace, "^C11$", "C11 early precursor spleen"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C10$", "C10 late precursor spleen"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C5$", "C5 naive CD4+ colon"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C4$", "C4 naive CD4+ spleen/fat"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C2$", "C2 naive CD4+ spleen"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C3$", "C3 naive CD4+ spleen"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C19$", "C19 Th1 skin"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C17$", "C17 Th1 colon"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C16$", "C16 Th1 spleen/fat"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C20$", "C20 Th17 skin"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C18$", "C18 Th2 colon"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C1$", "C1 undefined"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C6$", "C6 undefined"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C7$", "C7 undefined"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C9$", "C9 undefined"))

clusters_vector = as.vector(clusters$V1)
proj@cellColData$Clusters2 = clusters_vector
```

```{r celltype annotation, message=FALSE, warning=FALSE, eval=FALSE}
# Add celltype annotation to AchRProject
clusters = as.data.frame(proj@cellColData$Clusters)
colnames(clusters) = "V1"

clusters = clusters %>% mutate(across("V1", str_replace, "^C8$", "tisTreg")) 
clusters = clusters %>% mutate(across("V1", str_replace, "^C15$", "tisTreg"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C14$", "tisTreg"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C12$", "tisTreg"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C13$", "pTreg")) 
clusters = clusters %>% mutate(across("V1", str_replace, "^C11$", "tisTreg precursor"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C10$", "tisTreg precursor"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C5$", "naive CD4+"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C4$", "naive CD4+"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C2$", "naive CD4+"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C3$", "naive CD4+"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C19$", "Th1"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C17$", "Th1"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C16$", "Th1"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C20$", "Th17"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C18$", "Th2"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C1$", "undefined"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C6$", "undefined"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C7$", "undefined"))
clusters = clusters %>% mutate(across("V1", str_replace, "^C9$", "undefined"))

clusters_vector = as.vector(clusters$V1)
proj@cellColData$Celltype = clusters_vector
```

```{r plot cell type annotation, message=FALSE, warning=FALSE}
# Color by cell type
p_celltype = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "Celltype", 
  embedding = "UMAP", 
  size = 0.5
  )
p_celltype
```

```{r save ArchRProject after cluster anno, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
saveArchRProject(proj, 
  outputDirectory= "ArchRProject_final", 
  overwrite = TRUE,  
  load = FALSE
  )
```

## Reference data-based annotation using SingleR
```{r Reference data-based annotation using SingleR, warning=FALSE, message=FALSE, eval=FALSE}
# Reference data-based cell type annotation can be performed using SingleR.
# SingleR can be used with built-in reference datasets (for available datasets see
# (https://bioconductor.org/packages/release/bioc/vignettes/SingleR/inst/doc/SingleR.html))
# or using a custom reference dataset. 
# Reference dataset used: https://th-express.org

# Get gene score matrix from ArchRProject
GSM = getMatrixFromProject(
  ArchRProj = proj,
  useMatrix = "GeneScoreMatrix",
  useSeqnames = NULL,
  verbose = TRUE,
  binarize = FALSE,
  threads = getArchRThreads(),
  logFile = createLogFile("getMatrixFromProject")
)

# Extract relevant matrix and add rownames
GSM_matrix = assay(GSM)
rownames(GSM_matrix) = GSM@elementMetadata$name

# Get cell annotation reference
ref_data_stubbington = readRDS("Data/SingleR_refdata/singler_ref_stubbington_CD4.RDS")

ref_data_immgen = ImmGenData()
ref_data_immgen = ref_data_immgen[, ref_data_immgen$label.main %in% c("T cells") ]

# Annotate
GSM_stubbington = SingleR(
  test = GSM_matrix, 
  assay.type.test = 1,
  ref = ref_data_stubbington,
  labels = ref_data_stubbington$label.fine
)

table(predicted=GSM_stubbington$labels)

GSM_immgen = SingleR(
  test = GSM_matrix, 
  assay.type.test = 1,
  ref = ref_data_immgen,
  labels = ref_data_immgen$label.fine
)

table(predicted=GSM_immgen$labels)

# Add annotations to ArchRProject
proj@cellColData$pred_celltype_stubbington = GSM_stubbington$labels
proj@cellColData$pred_celltype_immgen = GSM_immgen$labels
```

```{r plotting SingleR cluster annotation, warning=FALSE, message=FALSE}
# Overlay annotations on UMAP
p_stubbington = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "pred_celltype_stubbington", 
  embedding = "UMAP", 
  plotAs = "points", 
  size = 0.5
)
p_stubbington

p_immgen = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "pred_celltype_immgen", 
  embedding = "UMAP", 
  plotAs = "points", 
  size = 0.5
)
p_immgen
```

```{r save SingleR cluster annotation, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Save
plotPDF(p_stubbington, p_immgen,
  name = "SingleR_stubbington_immgen.pdf", 
  ArchRProj = proj, 
  addDOC = FALSE, 
  width = 5, 
  height = 5
)
```

# Pseudo-bulk replicates
```{r review clusters with cell numbers, message=FALSE, warning=FALSE}
# We have a look at the number of cells in the individual clusters
cM_doubfilter = confusionMatrix(paste0(proj$Clusters2), paste0(proj$Sample))
cM_doubfilter
```

```{r make pseudo-bulk replicates, message=FALSE, warning=FALSE, eval=FALSE}
# The key parameter here is groupBy, which defines the groups for which 
# pseudo-bulk replicates should be made
proj = addGroupCoverages(
  ArchRProj = proj,
  groupBy = "Clusters2",
  minCells = 40, 
  maxCells = 500, 
  minReplicates = 2,
  maxReplicates = 5, 
  sampleRatio = 0.8,
  force = TRUE
  )
```

```{r save ArchRProject after computing pseudobulk replicates, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
saveArchRProject(proj, 
  outputDirectory= "ArchRProject_final", 
  overwrite = TRUE,  
  load = FALSE
  )
```

# Peak-calling
```{r peak calling, eval=FALSE, message=FALSE, warning=FALSE, error=TRUE}
# You can use the following function to search the path to Macs2
# However, sometimes this might not work and you have to manually add the path 
# like it is shown in the second line. 
pathToMacs2 = findMacs2()
# If you manually add the path, you have to change this line!
pathToMacs2 = "../../../../Macs2/bin/macs2"

proj = addReproduciblePeakSet(
  ArchRProj = proj, 
  groupBy = "Clusters2", 
  pathToMacs2 = pathToMacs2
)
```

```{r add peak matrix to ArchRProject, message=FALSE, warning=FALSE, eval=FALSE}
# Add peak matrix to ArchRProject
proj = addPeakMatrix(proj)

# Confirm that peak matrix has been added
getAvailableMatrices(proj)
```

```{r save ArchRProject post peak-calling, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
saveArchRProject(proj, 
  outputDirectory= "ArchRProject_final", 
  overwrite = TRUE,  
  load = FALSE
  )
```

## Tissue signatures
### Import tissue signatures
```{r custom enrichment - tissue signatures, message=FALSE, warning=FALSE, eval=FALSE}
# Check for enrichment of Treg signatures in Clusters
# signatures from PMID:33789089

signature_names = c("early_progenitor_tisTreg","late_progenitor_tisTreg","tisTreg_core",
  "tisTreg_skin","tisTreg_VAT","tisTreg_colon")
granges_list = list()
n = 1

for (x in signature_names){
  filename = paste("Data/signatures/",x,".csv", sep = "")
  sig_csv = read.csv(filename, sep = "-", header = FALSE)
  sig_csv = data.frame(seqnames=sig_csv[,1], start=as.character(sig_csv[,2]), end=as.character(sig_csv[,3]))
  granges_list[[n]] = makeGRangesFromDataFrame(sig_csv, seqnames.field="seqnames", 
  start.field="start", end.field="end")
  n = n+1
}

early_progenitor_tisTreg_GR = granges_list[[1]]
late_progenitor_tisTreg_GR = granges_list[[2]]
tisTreg_core_GR = granges_list[[3]]
tisTreg_skin_GR = granges_list[[4]]
tisTreg_VAT_GR = granges_list[[5]]
tisTreg_colon_GR = granges_list[[6]]
```

### Calculate signatures scores
```{r calculate signatures scores, message=FALSE, warning=FALSE, eval=FALSE}
# Signature z-scores can be calculated via the 'addDeviationsMatrix' function 
# of the chromVAR package. This wrapper function accepts the ArchRProject, the 
# GRanges objects with my signatures as a list, and the name for the signature 
# set as a string. The output is the ArchRProject with the z-scores and 
# deviation-scores added as additional columns to cellColData (one column for 
# each signature).

archr_add_peak_signatures = function(project, signature_list, signature_name){
  #signature_list: list of granges
  #signature_name: name string for the set of signatures
    add_df_to_cellcoldata = function(pro, pheno_df, force=FALSE){
      stopifnot(identical(rownames(pro@cellColData), rownames(pheno_df)))
      cnames = colnames(pheno_df)
      for(i in 1:ncol(pheno_df)){
        pro = addCellColData(ArchRProj = pro, data=pheno_df[, i], name = cnames[i],
                         cells = rownames(pro@cellColData), force = force)
      }
      return(pro)
    }
 if(length(signature_list)<2){
    stop('Currently, only works if at least two signatures are provided')
  }
  for(i in seq_along(signature_list)){
    names(signature_list[[i]]) = NULL
  }
  project = addPeakAnnotations(ArchRProj = project, 
                             regions = signature_list,
                             name = signature_name,
                             force=T)

  method_use = "chromVAR" #does only work with fixed width peaks
  if(any(sapply(signature_list, function(x) length(unique(width(x)))) > 1)){
     method_use = 'ArchR'
  }
  
  project = addBgdPeaks(project, force = T, method=method_use)
  
  project = addDeviationsMatrix(
    ArchRProj = project, 
    peakAnnotation = signature_name,
    binarize=TRUE,
    bgdPeaks = getBgdPeaks(project, method = method_use),
    force = TRUE
  )
  
  dr_df = as.data.frame(project@cellColData)
  sig_se = getMatrixFromProject(project, paste0(signature_name, 'Matrix'))
  z_score_mat = t(assays(sig_se)[['z']])
  z_score_mat = z_score_mat[match(rownames(dr_df),rownames(z_score_mat)), ]
  colnames(z_score_mat) = paste0('z_', colnames(z_score_mat))
  stopifnot(identical(rownames(z_score_mat), rownames(dr_df)))
  dev_score_mat = t(assays(sig_se)[['deviations']])
  dev_score_mat = dev_score_mat[match(rownames(dr_df),rownames(dev_score_mat)), ]
  colnames(dev_score_mat) = paste0('dev_', colnames(dev_score_mat))
  stopifnot(identical(rownames(dev_score_mat), rownames(dr_df)))
  
  project = add_df_to_cellcoldata(project, z_score_mat, force=T)
  project = add_df_to_cellcoldata(project, dev_score_mat, force=T)
  return(project)
}

signature_list = list(early_progenitor_tisTreg_sig = early_progenitor_tisTreg_GR, 
  late_progenitor_tisTreg_sig = late_progenitor_tisTreg_GR, 
  tisTreg_core_sig = tisTreg_core_GR, 
  tisTreg_skin_sig = tisTreg_skin_GR, 
  tisTreg_VAT_sig = tisTreg_VAT_GR, 
  tisTreg_colon_sig = tisTreg_colon_GR 
)

proj = archr_add_peak_signatures(proj, signature_list, "signatures")
```

```{r save ArchRProject after adding z- and dev-scores for signatures, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
# Save ArchRProject
saveArchRProject(proj, 
  outputDirectory= "ArchRProject_final", 
  overwrite = TRUE,  
  load = FALSE
  )
```

```{r overlay signatures on UMAP embedding, message=FALSE, warning=FALSE}
# Overlay signatures on UMAP
p_early_progenitor_tisTreg_sig = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "z_early_progenitor_tisTreg_sig", 
  embedding = "UMAP",
  plotAs = "points",
  size = 0.5
  )

p_late_progenitor_tisTreg_sig = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "z_late_progenitor_tisTreg_sig", 
  embedding = "UMAP",
  plotAs = "points",
  size = 0.5
  )

p_tisTreg_core_sig = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "z_tisTreg_core_sig", 
  embedding = "UMAP",
  plotAs = "points",
  size = 0.5
  )

p_tisTreg_skin_sig = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "z_tisTreg_skin_sig", 
  embedding = "UMAP",
  plotAs = "points",
  size = 0.5
  )

p_tisTreg_VAT_sig = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "z_tisTreg_VAT_sig", 
  embedding = "UMAP",
  plotAs = "points",
  size = 0.5
  )

p_tisTreg_colon_sig = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "cellColData", 
  name = "z_tisTreg_colon_sig", 
  embedding = "UMAP",
  plotAs = "points",
  size = 0.5
  )

# Plot
ggAlignPlots(p_early_progenitor_tisTreg_sig, p_late_progenitor_tisTreg_sig, 
  p_tisTreg_core_sig, type = "h")
ggAlignPlots(p_tisTreg_skin_sig, p_tisTreg_VAT_sig, p_tisTreg_colon_sig, type = "h")
```

```{r save p_tisTreg_sig p_Tfh_sig p_tisCD8_sig, message=FALSE, warning=FALSE, echo=FALSE}
# Save:
plotPDF(p_early_progenitor_tisTreg_sig, p_late_progenitor_tisTreg_sig, p_tisTreg_core_sig, p_tisTreg_skin_sig, p_tisTreg_VAT_sig, p_tisTreg_colon_sig,
  name = "UMAP_signatures.pdf", 
  ArchRProj = proj, 
  addDOC = FALSE, 
  width = 5, 
  height = 5
  )
```

## Identifying marker peaks
### Identifying marker peaks grouped by clusters
```{r identifying marker peaks (grouped by clusters), message=FALSE, warning=FALSE, results='hide'}
# Identify marker peaks (differential peaks) between clusters (cell types):
markersPeaks = getMarkerFeatures(
  ArchRProj = proj, 
  useMatrix = "PeakMatrix", 
  groupBy = "Clusters2",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon"
)

# Retrieve particular slices of this SummarizedExperiment that we are interested
# in: The default behavior of this function is to return a list of DataFrame 
# objects, one for each cell group
markerList = getMarkers(markersPeaks, cutOff = "FDR <= 0.01 & Log2FC >= 1")

# Instead of a list of DataFrame objects, we can use getMarkers() to return a 
# GRangesList object by setting returnGR = TRUE
markerList2 = getMarkers(markersPeaks, cutOff = "FDR <= 0.01 & Log2FC >= 1", returnGR = TRUE)

# Plot marker peaks as heatmap
heatmapPeaks = plotMarkerHeatmap(
  seMarker = markersPeaks, 
  cutOff = "FDR <= 0.001 & Log2FC >= 3",
  transpose = FALSE,
)
```

```{r plot markerpeaks in heatmap, error=TRUE, message=FALSE, warning=FALSE}
# Plot this heatmap using draw()
heatmap_peaks = ComplexHeatmap::draw(heatmapPeaks, 
  heatmap_legend_side = "right"
)
```

```{r save marker peak heatmaps by clusters, message=FALSE, warning=FALSE, echo=FALSE}
# Save as PDF
plotPDF(heatmap_peaks, 
  name = "Peak-Marker-Heatmap", 
  width = 8, 
  height = 6, 
  ArchRProj = proj, 
  addDOC = FALSE
)
```

## Pairwise testing between groups
### Pairwise testing between tisTreg and Treg clusters
```{r pairwise testing between clusters, warning=FALSE, message=FALSE}
# Get differential peaks between tisTreg and Treg cell clusters
markerTest = getMarkerFeatures(
  ArchRProj = proj, 
  useMatrix = "PeakMatrix",
  groupBy = "Clusters",
  testMethod = "wilcoxon",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  useGroups = tisTreg_cluster,
  bgdGroups = tTreg_cluster
)

# Visualize as volcano plot:
pv2 = markerPlot(seMarker = markerTest,
  name = tisTreg_cluster, 
  cutOff = "FDR <= 0.1 & abs(Log2FC) >= 1", 
  plotAs = "Volcano"
)

pv2
```

```{r pairwise testing betw clusters - save pv, message=FALSE, warning=FALSE, echo=FALSE}
# save as PDF:
plotPDF(pv2, 
  name = "tisTreg-vs-Treg-Markers-Volcano", 
  width = 5, 
  height = 5, 
  ArchRProj = proj, 
  addDOC = FALSE
)
```

# Motif enrichment
## Add motif annotations
```{r add MotifMatrix, warning=FALSE, message=FALSE, eval=FALSE}
# Look for motifs that are enriched in peaks that are up or down in various cell
# types
# We must first add these motif annotations to our ArchRProject; this 
# effectively creates a binary matrix where the presence of a motif in each peak
# is indicated numerically
proj = addMotifAnnotations(ArchRProj = proj, motifSet = "cisbp", name = "Motif")
```

```{r save ArchRProject after adding motif annotations, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
# Save ArchRProject
saveArchRProject(proj, 
  outputDirectory= "ArchRProject_final", 
  overwrite = FALSE,  
  load=F
  )
```

## Motif enrichment in marker peaks
### Motif enrichment in marker peaks grouped by clusters
```{r motif enrichment by clusters, warning=FALSE, message=FALSE, error=TRUE}
# We perform motif enrichment on our marker peaks 
# In this dataset we used more strict thresholds for the FDR and Log2FC 
# because of the size of the data; otherwise plots could not be generated.
enrichMotifs = peakAnnoEnrichment(
  seMarker = markersPeaks, 
  ArchRProj = proj,
  peakAnnotation = "Motif",
  cutOff = "FDR <= 0.001 & Log2FC >= 2"
)

# Plot these motif enrichments across all cell groups
heatmapEM = plotEnrichHeatmap(enrichMotifs, 
  n = 10, 
  transpose = TRUE
)

# Visualize
row_order = c("C5 naive CD4+ colon", "C4 naive CD4+ spleen/fat", "C2 naive CD4+ spleen", 
  "C3 naive CD4+ spleen", "C11 early precursor spleen", "C10 late precursor spleen", 
  "C8 tisTreg spleen", "C15 tisTreg fat", "C14 tisTreg skin", "C12 tisTreg colon", 
  "C13 pTreg colon", "C20 Th17 skin", "C16 Th1 spleen/fat", "C17 Th1 colon", "C19 Th1 skin", 
  "C18 Th2 colon", "C1 undefined", "C6 undefined", "C7 undefined", "C9 undefined")

heatmapEM2 = ComplexHeatmap::draw(heatmapEM, 
  heatmap_legend_side = "bot", 
  annotation_legend_side = "bot",
  row_order = row_order
)
```

```{r save motif enrichment heatmap by clusters, message=FALSE, warning=FALSE, echo=FALSE, error=TRUE}
# Save as PDF
plotPDF(heatmapEM2, 
  name = "Motifs-Enriched-Marker-Heatmap", 
  width = 8, 
  height = 6, 
  ArchRProj = proj, 
  addDOC = FALSE
)
```

## Motif enrichment in differential peaks
### Motif enrichment in differential peaks between tisTreg and Treg clusters
```{r motif enrichment in differential peaks between clusters, warning=FALSE, message=FALSE, error=TRUE}
# Define the set of significantly differential peaks that we are interested in 
# Create SummarizedExperiment object
motifsUp = peakAnnoEnrichment(
  seMarker = markerTest,
  ArchRProj = proj,
  peakAnnotation = "Motif", 
  cutOff = "FDR <= 0.1 & Log2FC >= 0.5"
)

# Prepare data for plotting with ggplot
# Create a simplified data.frame object containing the motif names, the corrected 
# p-values, and the significance rank
df_up = data.frame(TF = rownames(motifsUp), mlog10Padj = assay(motifsUp)[,1])
df_up = df_up[order(df_up$mlog10Padj, decreasing = TRUE),]
df_up$rank = seq_len(nrow(df_up))

# Plot rank-sorted TF motifs and color them by significance of their enrichment
ggUp = ggplot(df_up, aes(rank, mlog10Padj, color = mlog10Padj)) + 
  geom_point(size = 1) +
  ggrepel::geom_label_repel(
    data = df_up[rev(seq_len(30)), ], aes(x = rank, y = mlog10Padj, label = TF), 
    size = 1.5,
    nudge_x = 2,
    color = "black"
  ) + theme_ArchR() + 
  ylab("-log10(P-adj) Motif Enrichment") + 
  xlab("Rank Sorted TFs Enriched") +
  scale_color_gradientn(colors = paletteContinuous(set = "comet")
)
ggUp

# Perform the same analyses for the peaks that are more accessible in the Treg 
# compared to tisTreg by using peaks with Log2FC <= -0.5
motifsDo = peakAnnoEnrichment(
  seMarker = markerTest,
  ArchRProj = proj,
  peakAnnotation = "Motif",
  cutOff = "FDR <= 0.1 & Log2FC <= -0.5"
)

df_do = data.frame(TF = rownames(motifsDo), mlog10Padj = assay(motifsDo)[,1])
df_do = df_do[order(df_do$mlog10Padj, decreasing = TRUE),]
df_do$rank = seq_len(nrow(df_do))

ggDo = ggplot(df_do, aes(rank, mlog10Padj, color = mlog10Padj)) + 
  geom_point(size = 1) +
  ggrepel::geom_label_repel(
    data = df_do[rev(seq_len(30)), ], aes(x = rank, y = mlog10Padj, label = TF), 
    size = 1.5,
    nudge_x = 2,
    color = "black"
  ) + theme_ArchR() + 
  ylab("-log10(FDR) Motif Enrichment") +
  xlab("Rank Sorted TFs Enriched") +
  scale_color_gradientn(colors = paletteContinuous(set = "comet")
)
ggDo
```

```{r save ggUp ggDo, message=FALSE, warning=FALSE, echo=FALSE, error=TRUE}
# save as PDF:
plotPDF(ggUp, ggDo, 
  name = "tisTreg-vs-Treg-Motifs-Enriched", 
  width = 5, 
  height = 5, 
  ArchRProj = proj, 
  addDOC = FALSE
)
```

# ChromVAR deviations enrichment
ChromVAR is designed for computing accessibility scores for transcription factor motifs on a **per-cell basis** from sparse chromatin accessibility data

## Add motif deviations
```{r add background peaks and motif deviations matrix, message=FALSE, warning=FALSE, eval=FALSE}
# Add a set of background peaks; sample peaks based on similarity in GC-content 
# and nFrags across all samples using the Mahalanobis distance
proj = addBgdPeaks(proj, force = T)

# Compute per-cell deviations across all of our motif annotations
proj = addDeviationsMatrix(
  ArchRProj = proj,
  peakAnnotation = "Motif",
  force = TRUE
)
```

```{r save ArchRProject after calculating motif deviations, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
# Save ArchRProject
saveArchRProject(proj, 
  outputDirectory= "ArchRProject_final", 
  overwrite = T,  
  load=F
  )
```

```{r motif deviations, warning=FALSE, message=FALSE}
# Access deviations
# If we want this function to return a ggplot object, we set plot = TRUE 
# otherwise, this function would return the DataFrame object.
plotVarDev = getVarDeviations(proj, name = "MotifMatrix", plot = TRUE)

# plot variable deviations:
plotVarDev
```

```{r save plotVarDev, message=FALSE, warning=FALSE, echo=FALSE}
# Save plot as PDF
plotPDF(plotVarDev, 
  name = "Variable-Motif-Deviation-Scores", 
  width = 5, 
  height = 5, 
  ArchRProj = proj, 
  addDOC = FALSE
)
```

## Plot the distribution of motifs in the clusters
```{r subset motifs for downstream analysis, warning=FALSE, message=FALSE, error=TRUE}
# If we want to extract a subset of motifs for downstream analysis,
# we can do this using the getFeatures() function.
motifs = c("Batf")
markerMotifs = getFeatures(proj, select = paste(motifs, collapse="|"), useMatrix = "MotifMatrix")
markerMotifs

# Exclude Batf3
markerMotifs = grep("z:", markerMotifs, value = TRUE)
markerMotifs = markerMotifs[markerMotifs %ni% "z:Batf3_789"]
markerMotifs

# We supply the impute weights that we calculated previously during our gene 
# score analyses (impute weights allow us to smooth the signal across nearby 
# cells, helpful in the context of our sparse scATAC-seq data).

# First, re-run addImputeWeights (subset cells from the original imputation)
proj = addImputeWeights(proj)

# Then, plot distribution of chromVAR deviation scores
p_zscore = plotGroups(ArchRProj = proj, 
  groupBy = "Clusters", 
  colorBy = "MotifMatrix", 
  name = markerMotifs,
  imputeWeights = getImputeWeights(proj)
)

p_zscore
```

```{r save p_zscore, message=FALSE, warning=FALSE, echo=FALSE}
# Save plot as PDF
plotPDF(p_zscore, 
  name = "Plot-Groups-Deviations-w-Imputation", 
  width = 5, 
  height = 5, 
  ArchRProj = proj, 
  addDOC = FALSE
)
```

## Overlay motif z-score and corresponding gene score on UMAP
```{r overlay z-scores on UMAP embedding, message=FALSE, warning=FALSE}
# Instead of looking at the distributions of these z-scores, we can overlay the 
# z-scores on our UMAP embedding as we’ve done previously for gene scores.
# Overlay z-scores
p_zscore_UMAP_Batf = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "MotifMatrix", 
  name = "z:Batf_790", 
  embedding = "UMAP",
  plotAs = "points",
  imputeWeights = getImputeWeights(proj),
  size = 0.5
)

p_gene_exp_UMAP_Batf = plotEmbedding(
  ArchRProj = proj, 
  colorBy = "GeneScoreMatrix", 
  name = "Batf", 
  embedding = "UMAP",
  plotAs = "points",
  imputeWeights = getImputeWeights(proj),
  size = 0.5
)

plot_grid(p_zscore_UMAP_Batf, p_gene_exp_UMAP_Batf, align = "h", ncol = 2)
```

```{r save UMAP Batf z-score and gene score overlay, warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE}
# Save plot as PDF
plotPDF(p_zscore_UMAP_Batf, p_gene_exp_UMAP_Batf,
        name = "UMAP_Batf_motif-z-score_genescore.pdf", 
        ArchRProj = proj, 
        addDOC = FALSE, width = 5, height = 5)
```

# Motif footprinting
```{r motif footprinting, message=FALSE, warning=FALSE, error=TRUE}
# Obtain the positions of the relevant motifs
motifPositions = getPositions(proj, name = "Motif") 
motifPositions

# This creates a GRangesList object where each TF motif is represented by a 
# separate GRanges object. We can subset this GRangesList to a few TF motifs 
# that we are interested in.
motifs_fp = c("Batf", "Tbx21", "Gata3", "Rorc", "Rora", "Foxp3", "Ikzf2")
markerMotifs_fp = unlist(lapply(motifs_fp, function(x) grep(x, names(motifPositions), value = TRUE)))
markerMotifs_fp

# Exclude Batf3
markerMotifs_fp = markerMotifs_fp[markerMotifs_fp %ni% "Batf3_789"]
markerMotifs_fp

# To accurately profile TF footprints, a large number of reads is required. 
# Therefore we will use the pseudobulk data stored as group coverages calculated above.

# Compute footprints for the subset of marker motifs defined above
seFoot = getFootprints(
  ArchRProj = proj, 
  positions = motifPositions[markerMotifs_fp], 
  groupBy = "Clusters2"
)
```

## Motif footprinting in Treg cells
```{r subset project to Tregs, warning=FALSE, message=FALSE, eval=FALSE}
# Subset proj (with all matrices added) to only Treg cells
proj_treg = proj[(proj@cellColData@listData[["Clusters"]] %in% c(tisTreg_cluster,tTreg_cluster)), ] 
```

```{r save ArchRProject proj_treg, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
saveArchRProject(proj_treg, 
  outputDirectory= "ArchRProject_Treg", 
  overwrite = TRUE,  
  load = FALSE
  )
```

```{r LSI, clustering, UMAP, message=FALSE, warning=FALSE, eval=FALSE}
# LSI dimensional reduction
proj_treg = addIterativeLSI(
  ArchRProj = proj_treg,
  useMatrix = "TileMatrix",
  name = "IterativeLSI",
  iterations = 2, 
  clusterParams = list( 
    resolution = c(0.2), 
    sampleCells = 10000, 
    n.start = 10 
  ),
  varFeatures = 25000, 
  dimsToUse = 1:30,
  force = TRUE
)

# Clustering
proj_treg = addClusters(
  input = proj_treg,
  reducedDims = "IterativeLSI",
  method = "Seurat",
  name = "Clusters",
  resolution = 0.8, 
  force = TRUE
)

# Add UMAP embedding
proj_treg = addUMAP(
  ArchRProj = proj_treg, 
  reducedDims = "IterativeLSI", 
  name = "UMAP", 
  nNeighbors = 30, 
  minDist = 0.5, 
  metric = "cosine", 
  force = TRUE
)
```

```{r add group coverages, message=FALSE, warning=FALSE, eval=FALSE}
# The key parameter here is groupBy, which defines the groups for which 
# pseudo-bulk replicates should be made.
proj_treg = addGroupCoverages(
  ArchRProj = proj_treg,
  groupBy = "Clusters2",
  minCells = 40, 
  maxCells = 500, #standard
  minReplicates = 2, #standard
  maxReplicates = 5, #standard
  sampleRatio = 0.8,
  force = TRUE
  )
```

```{r save ArchRProject proj_treg 2, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
saveArchRProject(proj_treg, 
  outputDirectory= "ArchRProject_Treg", 
  overwrite = TRUE,  
  load = FALSE
  )
```

```{r load ArchRProject proj_treg, message=FALSE, warning=FALSE, include=FALSE}
# Load ArchRProject
proj_treg = loadArchRProject(
  path = "ArchRProject_Treg", 
  force = TRUE, 
  showLogo = FALSE
  )
```

```{r motif footprinting treg, message=FALSE, warning=FALSE, error=TRUE}
# Obtain the positions of the relevant motifs
motifPositions = getPositions(proj_treg, name = "Motif") 
motifPositions

# This creates a GRangesList object where each TF motif is represented by a 
# separate GRanges object. We can subset this GRangesList to a few TF motifs 
# that we are interested in.
motifs_fp = c("Batf", "Tbx21", "Gata3", "Rorc", "Rora", "Foxp3", "Ikzf2")
markerMotifs_fp = unlist(lapply(motifs_fp, function(x) grep(x, names(motifPositions), value = TRUE)))
markerMotifs_fp

# Exclude Batf3
markerMotifs_fp = markerMotifs_fp[markerMotifs_fp %ni% "Batf3_789"]
markerMotifs_fp

# To accurately profile TF footprints, a large number of reads is required. 
# Therefore we will use the pseudobulk data stored as group coverages calculated above.

# Compute footprints for the subset of marker motifs defined above
seFoot_treg = getFootprints(
  ArchRProj = proj_treg, 
  positions = motifPositions[markerMotifs_fp], 
  groupBy = "Clusters2"
)
```

## Normalization of footprints
### Normalization by subtraction of the Tn5 bias
```{r normalization by subtraction, message=FALSE, warning=FALSE, error=TRUE}
plotFootprints(
  seFoot = seFoot,
  ArchRProj = proj, 
  normMethod = "Subtract",
  plotName = "Footprints-Subtract-Bias",
  addDOC = FALSE,
  smoothWindow = 5
)
```

```{r normalization by subtraction 2, message=FALSE, warning=FALSE, error=TRUE}
plotFootprints(
  seFoot = seFoot_treg,
  ArchRProj = proj_treg, 
  normMethod = "Subtract",
  plotName = "Footprints-Subtract-Bias_Treg",
  addDOC = FALSE,
  smoothWindow = 5
)
```

```{r}
knitr::include_graphics("./ArchRProject_Treg/Plots/Footprints-Subtract-Bias_Treg.pdf")
```

### Normalization by dividing by the Tn5 bias
```{r normalization by division, message=FALSE, warning=FALSE, error=TRUE}
plotFootprints(
  seFoot = seFoot,
  ArchRProj = proj, 
  normMethod = "Divide",
  plotName = "Footprints-Divide-Bias",
  addDOC = FALSE,
  smoothWindow = 5
)
```

```{r normalization by division 2, message=FALSE, warning=FALSE, error=TRUE}
plotFootprints(
  seFoot = seFoot_treg,
  ArchRProj = proj_treg, 
  normMethod = "Divide",
  plotName = "Footprints-Divide-Bias_Treg",
  addDOC = FALSE,
  smoothWindow = 5
)
```

```{r}
knitr::include_graphics("./ArchRProject_Treg/Plots/Footprints-Divide-Bias_Treg.pdf")
```

### Footprinting without normalization for the Tn5 bias
```{r without normalization, message=FALSE, warning=FALSE, error=TRUE}
plotFootprints(
  seFoot = seFoot,
  ArchRProj = proj, 
  normMethod = "None",
  plotName = "Footprints-No-Normalization",
  addDOC = FALSE,
  smoothWindow = 5
)
```

```{r without normalization 2, message=FALSE, warning=FALSE, error=TRUE}
plotFootprints(
  seFoot = seFoot_treg,
  ArchRProj = proj_treg, 
  normMethod = "None",
  plotName = "Footprints-No-Normalization_Treg",
  addDOC = FALSE,
  smoothWindow = 5
)
```

```{r}
knitr::include_graphics("./ArchRProject_Treg/Plots/Footprints-No-Normalization_Treg.pdf")
```

# Co-accessibility analysis
```{r co-accessibility analysis, message=FALSE, warning=FALSE, eval=FALSE}
# Co-accessibility is a correlation in accessibility between two peaks 
# across many single cells. Calculate co-accessibility
proj = addCoAccessibility(
    ArchRProj = proj,
    reducedDims = "IterativeLSI"
)

# Retrieve co-accessibility information via the getCoAccessibility() function
# Increase resolution to plot fewer co-accessibility interactions
# if returnLoops = FALSE, then the function will return a dataframe instead of 
# a loop track
cA = getCoAccessibility(
    ArchRProj = proj,
    corCutOff = 0.5,
    resolution = 10000, 
    returnLoops = TRUE 
)

# Plot browser tracks of co-accessibility for our marker genes
markerGenes = c("Cd4", "Il2ra", "Foxp3", "Batf", "Ccr8", "Ikzf2", "Rorc", "Tbx21", "Ifng", "Gata3", "Il2", "Sell")

p_cA = plotBrowserTrack(
    ArchRProj = proj, 
    groupBy = "Clusters", 
    geneSymbol = markerGenes, 
    upstream = 50000,
    downstream = 50000,
    loops = getCoAccessibility(proj)
)

# To plot the browser track select a specific marker gene using the $ accessor
grid::grid.newpage()
grid::grid.draw(p_cA$Batf)
```

```{r save p_cA, message=FALSE, warning=FALSE, results='hide', echo=FALSE, eval=FALSE}
# Save plot as PDF
plotPDF(plotList = p_cA, 
    name = "Plot-Tracks-Marker-Genes-with-CoAccessibility.pdf", 
    ArchRProj = proj, 
    addDOC = FALSE, 
    width = 5, 
    height = 5
)
```

# Trajectory analysis
## ArchR's implementation
```{r trajectory analysis, message=FALSE, warning=FALSE, results='hide', error=TRUE, eval=FALSE}
# Create user-defined trajectory backbone from early via late progenitors to
# tissue Treg cells in the respective non-lymphoid tissue
Treg_trajectory_VAT = c("C11", "C10", "C8", "C15")
Treg_trajectory_skin = c("C11", "C10", "C8", "C14")
Treg_trajectory_colon = c("C11", "C10", "C8", "C12")

# Create the trajectory
# This creates a new column in cellColData that stores the pseudo-time value for 
# each cell in the trajectory
proj = addTrajectory(
    ArchRProj = proj, 
    name = "Treg_trajectory_VAT", 
    groupBy = "Clusters",
    trajectory = Treg_trajectory_VAT, 
    embedding = "UMAP", 
    force = TRUE
)

proj = addTrajectory(
    ArchRProj = proj, 
    name = "Treg_trajectory_skin", 
    groupBy = "Clusters",
    trajectory = Treg_trajectory_skin, 
    embedding = "UMAP", 
    force = TRUE
)

proj = addTrajectory(
    ArchRProj = proj, 
    name = "Treg_trajectory_colon", 
    groupBy = "Clusters",
    trajectory = Treg_trajectory_colon, 
    embedding = "UMAP", 
    force = TRUE
)

# Exclude cells with NA values because these are not part of the trajectory
proj$Treg_trajectory_VAT[!is.na(proj$Treg_trajectory_VAT)]
proj$Treg_trajectory_skin[!is.na(proj$Treg_trajectory_skin)]
proj$Treg_trajectory_colon[!is.na(proj$Treg_trajectory_colon)]

# Save ArchRProject
saveArchRProject(proj, 
  outputDirectory= "ArchRProject_final", 
  overwrite = TRUE,  
  load = FALSE
  )
```

```{r plot trajectory, message=FALSE, warning=FALSE, error=TRUE}
# Plot trajectory
# This overlays the pseudo-time values on our UMAP embedding and displays an 
# arrow approximating the trajectory path from the spline-fit; cells that are 
# not part of the trajectory are coloured in gray.
Treg_trajectory_VAT_p = plotTrajectory(proj, 
    trajectory = "Treg_trajectory_VAT", 
    colorBy = "cellColData", 
    name = "Treg_trajectory_VAT",
    plotAs = "points"
)

Treg_trajectory_skin_p = plotTrajectory(proj, 
    trajectory = "Treg_trajectory_skin", 
    colorBy = "cellColData", 
    name = "Treg_trajectory_skin",
    plotAs = "points"
)

Treg_trajectory_colon_p = plotTrajectory(proj, 
    trajectory = "Treg_trajectory_colon", 
    colorBy = "cellColData", 
    name = "Treg_trajectory_colon",
    plotAs = "points"
)

ggAlignPlots(Treg_trajectory_VAT_p[[1]], Treg_trajectory_skin_p[[1]], Treg_trajectory_colon_p[[1]], type = "h")
```

```{r save trajectories, message=FALSE, warning=FALSE, echo=FALSE, error=TRUE}
# Save plot as PDF
plotPDF(Treg_trajectory_VAT_p, Treg_trajectory_skin_p, Treg_trajectory_colon_p,
  name = "Plot-Treg_trajectory-UMAP.pdf", 
  ArchRProj = proj, 
  addDOC = FALSE, 
  width = 5, 
  height = 5
)
```

### Pseudotime UMAPs
```{r Treg pseudo-time UMAPs, message=FALSE, warning=FALSE, error=TRUE}
# Overlay other features on the trajectory within our UMAP embedding
# (display specific features only within the cells that are relevant to our 
# trajectory)

# Overlay genescores on Treg trajectory
# Batf
Treg_traj_VAT_p_Batf = plotTrajectory(proj, 
  trajectory = "Treg_trajectory_VAT", 
  colorBy = "GeneScoreMatrix", 
  name = "Batf", 
  continuousSet = "horizonExtra",
  plotAs = "points"
)

Treg_traj_skin_p_Batf = plotTrajectory(proj, 
  trajectory = "Treg_trajectory_skin", 
  colorBy = "GeneScoreMatrix", 
  name = "Batf", 
  continuousSet = "horizonExtra",
  plotAs = "points"
)

Treg_traj_colon_p_Batf = plotTrajectory(proj, 
  trajectory = "Treg_trajectory_colon", 
  colorBy = "GeneScoreMatrix", 
  name = "Batf", 
  continuousSet = "horizonExtra",
  plotAs = "points"
)

# Overlay motif scores on Treg trajectory
# Batf
Treg_traj_VAT_p_Batf_motif = plotTrajectory(proj, 
  trajectory = "Treg_trajectory_VAT", 
  colorBy = "MotifMatrix", 
  name = "z:Batf_790", 
  continuousSet = "solarExtra",
  plotAs = "points"
)

Treg_traj_skin_p_Batf_motif = plotTrajectory(proj, 
  trajectory = "Treg_trajectory_skin", 
  colorBy = "MotifMatrix", 
  name = "z:Batf_790", 
  continuousSet = "solarExtra",
  plotAs = "points"
)

Treg_traj_colon_p_Batf_motif = plotTrajectory(proj, 
  trajectory = "Treg_trajectory_colon", 
  colorBy = "MotifMatrix", 
  name = "z:Batf_790", 
  continuousSet = "solarExtra",
  plotAs = "points"
)

# Plot
ggAlignPlots(Treg_trajectory_VAT_p[[1]], Treg_traj_VAT_p_Batf[[2]], type = "h", sizes = c(1,2))
ggAlignPlots(Treg_trajectory_skin_p[[1]], Treg_traj_skin_p_Batf[[2]], type = "h", sizes = c(1,2))
ggAlignPlots(Treg_trajectory_colon_p[[1]], Treg_traj_colon_p_Batf[[2]], type = "h", sizes = c(1,2))

ggAlignPlots(Treg_trajectory_VAT_p[[1]], Treg_traj_VAT_p_Batf_motif[[2]], type = "h", sizes = c(1,2))
ggAlignPlots(Treg_trajectory_skin_p[[1]], Treg_traj_skin_p_Batf_motif[[2]], type = "h", sizes = c(1,2))
ggAlignPlots(Treg_trajectory_colon_p[[1]], Treg_traj_colon_p_Batf_motif[[2]], type = "h", sizes = c(1,2))
```

```{r save Treg trajectory UMAPs, message=FALSE, warning=FALSE, echo=FALSE, error=TRUE}
# Save plots as PDF
plotPDF(Treg_traj_VAT_p_Batf, Treg_traj_skin_p_Batf, Treg_traj_colon_p_Batf, Treg_traj_VAT_p_Batf_motif, Treg_traj_skin_p_Batf_motif, Treg_traj_colon_p_Batf_motif, 
  name = "Plot-Treg_trajectory_genes-UMAP.pdf", 
  ArchRProj = proj, 
  addDOC = FALSE, 
  width = 5, 
  height = 5
)
```

### Pseudo-Time Heatmaps
```{r Treg pseudo-time heatmaps, message=FALSE, warning=FALSE, error=TRUE}
# Visualize changes in features (MotifMatrix, GeneScoreMatrix) across 
# pseudo-time using heatmaps. 
# varCutOff (variance quantile cut-off) can be adjusted to set the top variable
# features across the trajectory

# Treg pseudotime MotifMatrix:
Treg_trajMM = getTrajectory(ArchRProj = proj, 
  name = "Treg_trajectory_VAT", 
  useMatrix = "MotifMatrix", 
  log2Norm = FALSE
)

# Plot:
Treg_trajMM_p = plotTrajectoryHeatmap(Treg_trajMM, 
  pal = paletteContinuous(set = "solarExtra"),
  varCutOff = 0.96
) 
Treg_trajMM_p


# Treg pseudotime GeneScoreMatrix:
Treg_trajGSM = getTrajectory(ArchRProj = proj, 
  name = "Treg_trajectory_VAT", 
  useMatrix = "GeneScoreMatrix", 
  log2Norm = FALSE
)

# Plot:
Treg_trajGSM_p = plotTrajectoryHeatmap(Treg_trajGSM, 
  pal = paletteContinuous(set = "horizonExtra"),
  varCutOff = 0.9985
) 
Treg_trajGSM_p 
```

```{r save Treg trajectory heatmaps, message=FALSE, warning=FALSE, echo=FALSE, error=TRUE}
# Save:
plotPDF(Treg_trajMM_p, Treg_trajGSM_p, 
  name = "Plot-Treg_VAT_trajectory_heatmap.pdf", 
  ArchRProj = proj, 
  addDOC = FALSE, 
  width = 5, 
  height = 5
)
```

```{r save ArchRProject proj, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
saveArchRProject(proj, 
  outputDirectory= "ArchRProject_final", 
  overwrite = TRUE,  
  load = FALSE
  )
```

```{r session info}
sessionInfo()
```